Towards Practical First-Order Model Counting

### 2025-01-16 doc/paper/paper.tex
Add proofs

\sisetup{group-separator = {,}}

@@

% TODO: the technical appendix should follow the same template (probably just
% integrate it into the main paper)

% TODO (for the entire paper):
% * 15 pages excluding references
% * draw a clear line between prior work and new developments

@@

It can be shown that the base cases identified by \FindBaseCases are sufficient
for the algorithm to terminate.\footnote{Note that characterising the
  fine-grained complexity of the solutions found by \Cranetwo{} or other FOMC
  algorithms is an emerging area of research. These questions have been
  partially addressed in previous
  work~\cite{DBLP:conf/kr/DilkasB23,tóth2024complexityweightedfirstordermodel}
  and are orthogonal to the goals of this section.}

\begin{theorem}[Termination]\label{thm:halting}
  Let $\mathcal{E}$ represent the equations returned by \CompileWithBaseCases.
  Let $f$ be an $n$-ary function in $\mathcal{E}$ and
  $\mathbf{x} \in \mathbb{N}_{0}^{n}$. Then the evaluation of $f(\mathbf{x})$
  terminates.
\end{theorem}

We prove \cref{thm:halting} using double induction. First, we apply induction to
the number of functions in $\mathcal{E}$. Then, we use induction on the arity of
the `last' function in $\mathcal{E}$ according to some topological ordering. For
the detailed proof, please refer to the technical appendix.

>>>>

\sisetup{group-separator = {,}}
\newtheorem{assumption}[theorem]{Assumption}

@@

% TODO (for the entire paper):
% * 15 pages excluding references

@@

It can be shown that the base cases identified by \FindBaseCases are sufficient
for the algorithm to terminate.\footnote{Note that characterising the
  fine-grained complexity of the solutions found by \Cranetwo{} or other FOMC
  algorithms is an emerging area of research. These questions have been
  partially addressed in previous
  work~\cite{DBLP:conf/kr/DilkasB23,tóth2024complexityweightedfirstordermodel}
  and are orthogonal to the goals of this section.} For the remainder of this
section, let $\mathcal{E}$ denote the equations returned by
\CompileWithBaseCases.

\begin{theorem}[Termination]\label{thm:halting}
  Let $f$ be an $n$-ary function in $\mathcal{E}$ and
  $\mathbf{x} \in \mathbb{N}_{0}^{n}$. Then the evaluation of $f(\mathbf{x})$
  terminates.
\end{theorem}

We prove \cref{thm:halting} using double induction. First, we apply induction to
the number of functions in $\mathcal{E}$. Then, we use induction on the arity of
the `last' function in $\mathcal{E}$ according to some topological ordering. We
begin with a few observations that stem from
previous~\cite{DBLP:conf/kr/DilkasB23,DBLP:conf/ijcai/BroeckTMDR11} and this
work. Let $\mathcal{E}$ represent the equations returned by
\CompileWithBaseCases.

\begin{observation}\label{assumption1}
  For each function $f$, there is precisely one equation $e \in \mathcal{E}$
  with $f(\mathbf{x})$ on the left-hand side where all $x_{i}$'s are variables
  (i.e., $e$ is not a base case). We refer to $e$ as the \emph{definition} of
  $f$.
\end{observation}

\begin{observation}\label{assumption2}
  There is a \emph{topological ordering} of all functions ${(f_{i})}_{i}$ in
  $\mathcal{E}$ such that equations in $\mathcal{E}$ with $f_{i}$ on the
  left-hand side do not contain function calls to $f_{j}$ with $j > i$. This
  condition prevents mutual recursion and other cyclic scenarios.
\end{observation}

\begin{observation}\label{assumption3}
  For every equation $(f(\mathbf{x}) = \expr) \in \mathcal{E}$, the evaluation
  of $\expr$ terminates when provided with the values of all relevant function
  calls.
\end{observation}

\begin{corollary}\label{fact}
  If $f$ is a non-recursive function with no function calls on the right-hand
  side of its definition, then the evaluation of any function call
  $f(\mathbf{x})$ terminates.
\end{corollary}

\begin{observation}\label{fact2}
  For any equation $f(\mathbf{x}) = \expr{}$, if $\mathbf{x}$ contains only
  constants, then $\expr{}$ cannot include any function calls to $f$.
\end{observation}

Additionally, we introduce an assumption about the structure of recursion.

\begin{assumption}\label{assumption4}
  For every equation $(f(\mathbf{x}) = \expr) \in \mathcal{E}$, every recursive
  function call $f(\mathbf{y}) \in \expr$ satisfies the
  following:
  \begin{itemize}
    \item Each $y_{i}$ is either $x_{i} - c_{i}$ or $c_{i}$ for some constant
          $c_{i}$.
    \item There exists $i$ such that $y_{i} = x_{i} - c_{i}$ for some
          $c_{i} > 0$.
  \end{itemize}
\end{assumption}

Finally, we assume a particular order of evaluation for function calls using the
equations in $\mathcal{E}$. Specifically, we assume that base cases are
considered before the recursive definition. The exact order in which base cases
are considered is immaterial.

\begin{assumption}
  When multiple equations in $\mathcal{E}$ match a function call
  $f(\mathbf{x})$, preference is given to an equation with the most constants on
  its left-hand side.
\end{assumption}

With the observations and assumptions mentioned above, we are ready to prove
\cref{thm:halting}. For readability, we divide the proof into several lemmas of
increasing generality.

\begin{lemma}\label{lemma:oneunary}
  Assume that $\mathcal{E}$ consists of just \emph{one unary} function $f$. Then
  the evaluation of a function call $f(x)$ terminates for any
  $x \in \mathbb{N}_{0}$.
\end{lemma}
\begin{proof}
  If $f(x)$ is captured by a base case, then its evaluation terminates by
  \cref{fact,fact2}. If $f$ is not recursive, the evaluation of
  $f(x)$ terminates by \cref{fact}.

  Otherwise, let $f(y)$ be an arbitrary function call on the right-hand side of
  the definition of $f(x)$. If $y$ is a constant, then there is a base case for
  $f(y)$. Otherwise, let $y = x - c$ for some $c > 0$. Then there exists
  $k \in \mathbb{N}_{0}$ such that $0 \le x - kc \le c-1$. So, after $k$
  iterations, the sequence of function calls $f(x), f(x-c), f(x-2c),\dots$ will
  be captured by the base case $f(x \mod c)$.
\end{proof}

\begin{lemma}\label{lemma:onefunction}
  Generalising \cref{lemma:oneunary}, let $\mathcal{E}$ be a set of equations
  for \emph{one} $n$-ary function $f$ for some $n \ge 1$. Then the evaluation of
  $f(\mathbf{x})$ terminates for any $\mathbf{x} \in \mathbb{N}_{0}^{n}$.
\end{lemma}
\begin{proof}
  If $f$ is non-recursive, the evaluation of $f(\mathbf{x})$ terminates by
  previous arguments. We proceed by induction on $n$, with the base case of
  $n=1$ handled by \cref{lemma:oneunary}. Assume that $n > 1$. Any base case of
  $f$ can be seen as a function of arity $n-1$, since one of the parameters is
  fixed. Thus, the evaluation of any base case terminates by the inductive
  hypothesis. It remains to show that the evaluation of the recursive equation
  for $f$ terminates, but that follows from \cref{assumption3}.
\end{proof}

\begin{proof}[Proof of \cref{thm:halting}]
  We proceed by induction on the number of functions $n$. The base case of $n=1$
  is handled by \cref{lemma:onefunction}. Let ${(f_{i})}_{i=1}^{n}$ be some
  topological ordering of these $n > 1$ functions. If $f = f_{j}$ for $j < n$,
  then the evaluation of $f(\mathbf{x})$ terminates by the inductive hypothesis
  since $f_{j}$ cannot call $f_{n}$ by \cref{assumption2}. Using the inductive
  hypothesis that all function calls to $f_{j}$ (with $j < n$) terminate, the
  proof proceeds similarly to the Proof of \cref{lemma:onefunction}.
\end{proof}
